---
title: "Week 7 - Time Series, Data Formats, Output Formats, Project Intro"
author: "Penelope Pooler Eisenbies"
date: "`r Sys.Date()`"
output: 
  html_document: 
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=TRUE}

# this line specifies options for default options for all R Chunks
knitr::opts_chunk$set(echo=T, 
                      highlight=T)

# suppress scientific notation
options(scipen=100,
        getSymbols.warning4.0 = FALSE)

# install helper package (pacman), if needed
if (!require("pacman")) install.packages("pacman", repos = "http://lib.stat.cmu.edu/R/CRAN/")

# install and load required packages
# pacman should be first package in parentheses and then list others
pacman::p_load(pacman, tidyverse, gridExtra, magrittr, lubridate, 
               knitr, kableExtra, tidyquant, highcharter, dygraphs, 
               htmlwidgets)

# verify packages (comment out in finished documents)
p_loaded()

```

---

### Quiz 1 

- Quiz 1 is now graded

  - 10%: Submitted Markdown File
  
  - 90%: Blackboard Answers + .csv files and .png file
  
- Please don't worry if you are not happy with your score
  
- Quiz 2 will be during Week 11 and will combine previous skills with material from weeks 6 through 10

  - Same format but may have more questions
  
  - More multi-step tasks

- If you have questions about your quiz, please let me know.

---

### BUA 455 Group Dashboard Project

#### Groups

- Complete HW 4 - Part 1 TODAY! (This should only take 5 min.)

  - **Note:** If you do not complete this Survey, I will not put you in a project group and you can not pass this class.
  
- Groups of 5 or 6 will be determined and posted (Hopefully by Friday)

- If you have a request to work with someone, include that information in your survey (Not required).

  - **Tomorrow (2/28) is the last day I will accept any group requests.**

  - I cannot guarantee that requests will be honored, but I will try.
  
<br>

#### Project Information Links

- [Project Description](https://docs.google.com/document/d/1U-DJ3yeHPpxcg1o12Cg2qc2Besb6Jw6UiyAo6gR4S2I/edit?usp=sharing)

- Project Ideas, Data Sources, etc. also available, and will be updated.

---

### Upcoming Dates

- **Groups assigned by Monday (3/4) at the latest**

- **Mon.  3/11 - Fri. 3/15 SPRING BREAK** 
  
- **Thu. 3/21 at 5:00 PM:** Draft Proposals Due (2 - 3 paragraphs) - NO GRACE PERIOD

- **Tue. 3/26: Proposal Meetings** 
  - Groups should come with questions and be prepared to answer my questions (10-15 min. per groups)
  
-  **Wed. 3/27:**  HW 5 - Part 1 Due

-  **Thu. 4/4:**  Quiz 2

-  **Tue. 4/9:**  Final Proposals Due

---

### Reminders about HW 4

- Chunk Headers  
  
  - In Chunk 6 (Part 5), the chunk header in the the template appears as follows:

    ```
    {r create and save plot, eval=F}

    ```
- The `eval=F` prevents this chunk from being evaluated when it is knit.

- It was included in the template because the original code provided was incomplete and incorrect and would cause errors when knit.

- You are asked to remove the comma and the text `eval=F`

- There are many other chunk header options.

- Two common ones that you will use in your project are `echo=F` and `include=F`

- SIDE NOTE: If two chunks are given the EXACT SAME name, e.g. `importing data`, document will not knit.

- **Markdown References**

  - [Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/main/rmarkdown-2.0.pdf)
  
  - [Reference Guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)

---

### Markdown Output Formats

- So far, all work in this course has been done in R Markdown and we have intentionally used the default output format, `.html`.

- Other common formats are Word documents (see example), PDF documents, Powerpoint Slides, and **dashboards**

#### Dashboard Templates in the `flexdashboard` package

- There are many types of dashboards. Here are [some example dashboards](https://rstudio.github.io/flexdashboard/articles/examples.html) created with `flexdashboard` and other packages in R

- For this course (which focuses on Data Management), we will use a excellent introductory option, [storyboard](https://beta.rstudioconnect.com/jjallaire/htmlwidgets-showcase-storyboard/htmlwidgets-showcase-storyboard.html)
  
  - This template simplifies the dashboard development process 
  
  - Can be used to present data and visualizations in a progression

- The Storyboard template is an R Markdown file with some very **specific pre-set features**

  - **These minor details are required** or the dashboard will not knit correctly.

  - HW 5 - Part 1 and HW 5 - Part 2 will be completed in a storyboard dashboard format to give you practice working with data in this way.

---

### Types of Time Series Data in R

- In recent weeks, we have worked with Box Office Mojo and Bureau of Labor Statistics Data

- These datasets all include a date variable and another quantitative variable that changes at each time period, e.g. 

  - movie gross (HW 3)
  - unemployment rate (HW 4)
  - import and export indices (HW 4)
  
- All of these datasets are considered time series but the data format in R is called a `tibble`.

- Two common data formats in R, `tibble` and `data.frame` are needed for creating ggplots of time series.
  `tibble` are the more modern format and are more compatible with `tidyverse` commands to manage data.
  
- Today, we'll discuss a third data format, `xts` 


---

### Importing Stock Data as `xts` using `tidyquant` Package

- [Yahoo Finance](https://finance.yahoo.com/), the Federal Reserve Bank, WSJ and others are excellent data sources that can be directly imported into R
  
  - default for `getsymbols` in the `tidyquant` package is Yahoo Finance
  
  - Data format is `xts` which we will cover today
  
```{r importing data from yahoo finance, results='hide'}

# download data from Netflix, Amazon, Disney
# time series starts day after from date specified
# time series ends day before to date specified
getSymbols("NFLX", from = "2014-01-01", to = "2023-09-30")
getSymbols("AMZN", from = "2014-01-01", to = "2023-09-30")
getSymbols("DIS", from = "2014-01-01", to = "2023-09-30")

```

---

### Example of Imported Yahoo Finance Data

By default, imported dates are assigned to stock symbol in ALL CAPS, e.g. **NFLX**

```{r display of imported yahoo finance data}

head(NFLX)

```

---

### Example of `hchart` for One Stock

`hchart` in the `highcharter` package is one way to plot `xts` data

```{r hchart of 1 stock, fig.dim=c(15,5)}
hchart(NFLX$NFLX.Adjusted, name="Adjusted", color="green") |>   # plot adjusted close
  hc_add_series(NFLX$NFLX.High, name="High" , color="blue") |>  # add daily high
  hc_add_series(NFLX$NFLX.Low, name="Low" , color="red")        # add daily low
```

---

### Example of `hchart` for Three Stocks

```{r  hchart of 3 stocks, fig.dim=c(15,5)}
hchart(AMZN$AMZN.Adjusted, name="AMZN", color="green") |>      # plot AMZN adjusted close
  hc_add_series(DIS$DIS.High, name="DIS" , color="blue") |>    # add DIS adjusted close
  hc_add_series(NFLX$NFLX.Low, name="NFLX" , color="red")      # add NFLX adjusted close

```

---

### R code for a Multi-Panel Composition of `hcharts`

- Stocks can be shown in separate plots that can be shown side by side or in one stacked column

- The command `hw_grid` is used to display them and `ncol` indicates how many columns.

```{r separate stock plots}
nflx_plt <- hchart(NFLX$NFLX.Adjusted, name="Adjusted", color="green") |>
  hc_add_series(NFLX$NFLX.High, name="High" , color="darkgreen") |>
  hc_add_series(NFLX$NFLX.Low, name="Low" , color="lightgreen")

amzn_plt <- hchart(AMZN$AMZN.Adjusted, name="Adjusted", color="blue") |>
  hc_add_series(AMZN$AMZN.High, name="High" , color="darkblue") |>
  hc_add_series(AMZN$AMZN.Low, name="Low" , color="lightblue")

dis_plt <- hchart(DIS$DIS.Adjusted, name="Adjusted", color="mediumpurple") |>
  hc_add_series(DIS$DIS.High, name="High" , color="purple4") |>
  hc_add_series(DIS$DIS.Low, name="Low" , color="plum")
```

---

### Display of Multi-Panel Composition of `hcharts`

```{r display of plots, fig.dim=c(15,6)}
hw_grid(nflx_plt, amzn_plt, dis_plt, ncol=3)
```

---


class: middle

### Week 7 In-class Exercises (Q1)

***PointSolutions Session ID: bua455f23***

In the example above, we use the `hw_grid` command to create multi-plot composition of hcharts.

Previously, we covered another command to do the same thing with non-interactive ggplots of `tibble` data.

What is that other command?  

Hints: This very useful command is in the `gridExtra` package which is loaded.

If `gridExtra` is loaded in R, start typing `grid` in the console, and the command and others will appear.

---


class: middle

### Week 7 In-class Exercises (Q2)

***PointSolutions Session ID: bua455f23***

1.  Use provided exampled of `getSymbols` code to write code to import the stock time series for Apple (`AAPL`)
    - Use these dates: from = "2015-01-01", to = "2023-10-01"
    
2.  Open the imported `xts` file by clicking on it in the `Global Environment`

3.  Sort the `AAPL.Adjusted` column by clicking on it.

4.  Answer **AAPL** Question:
    -   On what recent date, was AAPL at it's highest value?

```{r}

```



---

### `XTS` Data

#### More Information about `xts` 

- When these stock datasets are imported, they are in `xts` format.

- `xts` stands for **Extensible Time Series** which means they are self-aware.

- The key feature is that `date` is NOT a variable, but instead the dates become row IDs.

  - Any dataset with a `date` variable can be converted to an `xts` dataset.

   - Any `xts` dataset can be converted a tibble or data.frame (two common R data formats).
   
```{r examine xts data}
head(NFLX)
```

---

### Merging `xts` datasets using merge

- Converting xts to a tibble or dataframe (R data formats) is required if you want to create a ggplot or use other methods covered previously

- A good first step is to create a merged xts dataset of the variables you want.

```{r merge xts stock data }
# all merged xts datasets must have same date index
nflx_amzn_dis <- merge(NFLX$NFLX.Adjusted,
                       AMZN$AMZN.Adjusted,
                       DIS$DIS.Adjusted) 
head(nflx_amzn_dis)
```

---

### Converting `xts` datasets to tibble format

- There are a few ways to convert an xts to a tibble.

- In the code below I show the conversion and then I rename the the new date variable as `date`

```{r convert xts to tibble}
# converting data to a tibble requires a couple lines of code 
# I prefer to rename the index as date 
nflx_amzn_dis_tibble <- nflx_amzn_dis |> 
  fortify.zoo() |> as_tibble(.name_repair = "minimal") |>
  rename("date" = "Index") 
head(nflx_amzn_dis_tibble)

```

---

### Converting tibble format datasets to `xts`

- Any dataset with a date formatted variable can be converted to an `xts` dataset
- We can use `hchart` or `dygraph` (next topic) for any time series

```{r convert tibble to xts dataset and plot, fig.dim=c(15,4)}

exp_imp <- read_csv("export_import_tidy.csv", show_col_types=F)
exp_imp_xts <- xts(x=exp_imp[,2:3], order.by=exp_imp$date) # order.by must be a date variable

(exp_imp_hc <- hchart(exp_imp_xts$exp_indx, name="Export Price Index", color="blue") |>
  hc_add_series(exp_imp_xts$imp_indx, name="Import Price Index" , color="red"))
```

---

### Dygraphs - One Alternative to `hchart` for `XTS` data

#### Dygraph Info

- `dygraph` is A more flexible alternative to `hchart`.

- Both `hchart` and `dygraph` are useful and which you choose depends on your goal(s).

  - Many students find `dygraphs` useful for projects.
  - You can add reference lines and shading
  - Both `dygraph` and `hchart` allow viewer to interactively select date range
  
Here is the dataset we will use:

```{r create dataset for dygraphs}
three_stocks <- merge(AMZN$AMZN.Adjusted, DIS$DIS.Adjusted, NFLX$NFLX.Adjusted) 
names(three_stocks) <- c("AMZN.adj", "DIS.adj", "NFLX.adj")
head(three_stocks, 3) # print first three rows only
```

---

### Unformatted `dygraph`

This first plot is basic unformatted plot of three stocks with the range selector option.

```{r basic dygraph with range selector, fig.dim=c(15,4)}

(dy3 <- dygraph(three_stocks, main="Streaming Company Stock Trends") |>
  dySeries("AMZN.adj", label="AMZN", color= "green") |>
  dySeries("DIS.adj", label="DIS", color= "red") |>
  dySeries("NFLX.adj", label="NFLX", color= "blue") |>
  dyRangeSelector())

```


### `dygraph` with Some formatting

Two useful formatting options (shown below) to make the plot more readable are:
 - Removing the the grid lines 
 - Formatting the axis labels
```{r dygraph with axes labeled and gridlines removed, fig.dim=c(15,3.5)}
(dy3 <- dy3 |>
  dyAxis("y", label = "Adjusted Close", drawGrid = FALSE) |>
  dyAxis("x", label = "Date", drawGrid = FALSE))
```
 
---

### Adding Event Lines

Vertical lines can be added at specific dates and can be labeled and formatted.

```{r dygraph with event lines, fig.dim=c(15,4)}
(dy3 <- dy3 |>
  dyEvent("2020-3-12", label = "Theaters Closed", labelLoc = "bottom") |>
  dyEvent("2021-6-15", label = "Restrictions End", labelLoc = "bottom", strokePattern = "solid"))
```

---

### Adding Shaded Regions

Alternatively, it may be helpful to shade plot for a specific time range.

```{r dygraph with shaded time period, fig.dim=c(15,4)}
(dy3 <- dy3 |>
  dyShading(from = "2020-3-12", to = "2021-6-15", axis = "x", color = "lightgrey"))
```

---

### Review of `bls_tidy` Function for Labor Data

- Before using our function on new data, we **ALWAYS** examine the .csv files

- The number of rows to skip for these three labor datasets is **`11`**

```{r run bls_tidy and import labor data}

bls_tidy <- function(data_file, skip_num, var_name){
  read_csv(data_file, skip = skip_num, show_col_types = F) |> 
  pivot_longer(cols = Jan:Dec,                      
               names_to = "month", 
               values_to = "value") |>
  filter(!is.na(value)) |>                    
  rename({{var_name}} := "value")                             
}

labor_force <- bls_tidy("bls_civ_lf.csv", skip_num=11, var_name="lf")
unemp <- bls_tidy("bls_civ_unemp.csv", skip_num=11, var_name="unemp")
emp <- bls_tidy("bls_civ_emp.csv", skip_num=11, var_name="emp")

head(unemp)

```

---

### Joining More than Two Datasets

- Last Week and in HW 4 we covered joining **2** datasets.

- The commands we covered (there are 4) all have the same limitation: **datasets must be joined two at a time.**

```{r joining labor force data with pipes}
# with piping
lf_all <- labor_force |>
  full_join(emp) |>
  full_join(unemp) |>
  write_csv("labor_tidy.csv")
head(lf_all)

```

<br>

```{r joining labor force data without pipes}
# without piping
lf_all <- full_join(labor_force, emp) 
lf_all <- full_join(lf_all, unemp) 
head(lf_all)

```

---

### Creating a Date Variable and Modifying Data for a Plot (Review)

- Includes code that is similar to Parts 3 and 4 of HW 4.
- BONUS: Code modified to show how to get 'End of Month' (eom) date (**[Useful Link](https://www.statology.org/lubridate-first-last-day-of-month/)**)

```{r dates and data mod for plot}

lf_plt <- lf_all |>
  mutate(date_som = ym(paste(Year, month)),         # create som date var
         date = ceiling_date(date_som, "month")-1,  # create eom month date var
         empM = (emp/1000) |> round(2),             # convert counts to millions
         unempM = (unemp/1000) |> round(2)) |>
  select(date, empM, unempM) |>                     # select vars and reshape
  pivot_longer(cols=empM:unempM, names_to = "type", values_to = "count") |>
  mutate(type = factor(type,                        # create factor var for plot
                       levels = c("unempM", "empM"),
                       labels = c("Unemployed", "Employed"))) 

head(lf_plt, 4) # examine first 4 rows

```

---

### Creating a Polished Area Plot for Slides

- Useful for data that sum to a whole: **Employed + Unemployed = Total Labor Force**

```{r plot code for lf area plot}

lf_area_plt_slides <- lf_plt |>
  ggplot() +
  geom_area(aes(x=date, y=count, fill=type)) +
  theme_classic() +
  theme(legend.position="bottom") +
  scale_fill_manual(values=c("red", "blue")) + 
  scale_x_date(date_breaks = "year", date_labels = "%Y") +
  labs(x="Date", y = "Number of Peolple (Millions)", fill="",
       title="Total Labor Force: Employed and Unemployed ", 
       subtitle="Jan. 2012 - Sep. 2022",
       caption="Data Source:www.bls.gov") + 
  theme(plot.title = element_text(size = 20),                    
        plot.subtitle = element_text(size = 15),
        axis.title = element_text(size=18),
        axis.text = element_text(size=15),
        plot.caption = element_text(size = 10),
        legend.text = element_text(size = 12),
        panel.border = element_rect(colour = "lightgrey", fill=NA, size=2),
        plot.background = element_rect(colour = "darkgrey", fill=NA, size=2))

```

---

### Area Plot Formatted for Slides

```{r display of final area plot, echo=F, fig.dim=c(15,7)}

lf_area_plt_slides

```

---

### Creating an Area Plot for HTML, Documents and Export

- Additional formatting (shown above) can always be added

```{r simpler plot code for lf area plot }

lf_area_plt <- lf_plt |>
  ggplot() +
  geom_area(aes(x=date, y=count, fill=type)) +
  theme_classic() +
  theme(legend.position="bottom") +
  scale_fill_manual(values=c("red", "blue")) + 
  scale_x_date(date_breaks = "year", date_labels = "%Y") +
  labs(x="Date", y = "Number of Peolple (Millions)", fill="",
       title="Total Labor Force: Employed and Unemployed ", 
       subtitle="Jan. 2012 - Sep. 2022",
       caption="Data Source:www.bls.gov")

```

#### Using R code to export a plot or image 
- There are **MANY** other options besides `.png`
```{r export plot as png file}
png("labor_force_area_plot.png", height=480, width=720)
lf_area_plt
dev.off()

```

---

### Exported Plot

- Looks fine in HTML notes but not slides
- May be fine in Word Document or Dashboard
- If not, previous code shows options for formatting

```{r display of plot that was exported, fig.dim=c(15,6), echo=F}
lf_area_plt

```

---

### Week 7 In-class Exercise

In this exercise we will 

**`1.`** Import `labor_tidy.csv` and convert variables to millions and round to 2 decimal places and select two variables. (Review)

- OPTIONAL: use provided example to create an END of Month (eom) date variable and use that for exercise.
   
```{r import labor_tidy and modify vars}

labor_new <- read_csv("labor_tidy.csv") |>
  mutate(date = ym(paste(Year,month)),
         lfM = (lf/1000) |> round(2),
         empM = (emp/1000) |> round(2))|>
  select(date, lfM, empM)

```

---

**`2.`** Convert `labor_new` to an `xts` format, `labor_xts`

```{r convert labor_new to xts}
# labor_xts
```


**`3`**. Create a `hchart` with two variables
   - No additional formatting is expected (difficult in hchart)
   - Plot `lfM` and `empM` and save it as `labor_hc`
```{r create and display labor hchart}
# (labor_hc <- hchart())
```


**`4.`** Create a `dygraph` with same two variables from `hchart`
   - Include `dyRangeSelector`
   - Add axis labels to `dygraph` using `dyAxis`
   - Optional but not required: adding `dyEvent` lines or `dyShading`
   
```{r create and display labor dygraph}
# (labor_dy <- dygraph())
```


**`5.`** Show me your 2 plots by end of class on Thursday (if possible)
  
- Interactive plots can be exported as HTML files

- Blackboard does not let students submit .html files because of virus concerns.

Code to export a dygraph as an html file (`htmlwidgets` package required):

`saveWidget(widget = labor_dy, file="labor_dy.html")`
   
**`6.`** Final Step: Although we use R markdown files in this course, it is sometimes useful to save R code as an R Script.

In the R project `code_data_output` folder I have saved an R Script for your work.

- Copy and paste code into provided R Script and use save as to save the file with your name., e.g. `Week_7_In_Class_Penelope_Pooler.R`

- R Script should include: 

  - code I provided to import and modify data

  - tibble to xts conversion of labor dataset

  - hchart plot code

  - dygraph plot code

  - add comments to code using `#` so that you can understand it and use it later.

- Submit final script on Blackboard.  This will count towards your class particpation for Week 7.


---


### Additional Comments about R Markdown files and R Scripts
 

- Markdown files are 'smart': File is aware of it's location 

- R Scripts (older common file type) are useful BUT not aware of file location.
  - User must specify working directory
  
  - The script I provided is saved to your working directory
  
- To check working directory:  `getwd()`

- To set working directory to code_data_output folder: (for working in an R Script)
  - Click Session > Set Working Directory > To Source File Location

NOTES: 

- R users and developers do not recommend setting working directories

  - Have to be changed in the R code for each computer
  
  - Hampers collaboration and code execution

- When possible use R Projects, code_data_output folders, and R Markdown files 

  - No need to set working directory
  
  - As you advance, you may use separate folders for code, data, and output within projects.
  
---

### **Key Points from Week 7**

**Time Series Data**

- Importing stock data from Yahoo Finance as `xts` 

- Converting between `xts` and `tibble`

- Plotting options include area plots, hcharts and dygraphs

- `dygraphs` and `hcharts` are useful tools for understanding, managing, and curating time series data.

- HW 4 is due Wednesday, 2/28

  - Grace period in effect.  TA and I are available to assist if you have questions.
  
<br>

You may submit an 'Engagement Question' about each lecture until midnight on the day of the lecture. **A minimum of four submissions are required during the semester.**

